---
title: "gap-minder bubble chart"
author: "Maeva"
date: 2017-03-31
output: 
  html_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center")
```

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(plotly)
```

# Introduction

[Gapminder]("http://www.gapminder.org") is a foundation that maintains a database of global socio-economic numbers about each country across time, to promote data-driven sustainable development worlwide. 

One of their iconic visualization shows [how life expectancy relates to gdp per capita across time]("http://www.gapminder.org/tools/chart-type=bubbles"). 

I am trying to recreate this vis using `ggplot2` (for the static version), and `plotly` (for the interactive version).

# Data

The data is maintained and available on GitHub.

```{r}
# URLs
life_expectancy_url <- 
  "https://raw.githubusercontent.com/open-numbers/ddf--gapminder--life_expectancy/master/ddf--datapoints--life_expectancy--by--geo--time.csv"

gdp_capita_year_url <- 
  "https://raw.githubusercontent.com/open-numbers/ddf--gapminder--gdp_per_capita_cppp/master/ddf--datapoints--gdp_per_capita_cppp--by--geo--time.csv"

geo_url <- 
  "https://raw.githubusercontent.com/open-numbers/ddf--gapminder--gdp_per_capita_cppp/master/ddf--entities--geo.csv"

continent_url <- 
  "https://docs.google.com/spreadsheets/d/1I9Bt2jnafZKGyWDry4RAnZ_DopLYw9t_SjIiAZwZaGY/pub?gid=1"

population_url <- 
  "https://raw.githubusercontent.com/open-numbers/ddf--gapminder--population/master/ddf--datapoints--population--by--country--year.csv"
```

# Loading into tibbles

## Life expectancy
```{r}
life_expectancy <- 
  life_expectancy_url %>% 
  read_csv()

if (sum(is.na(life_expectancy)) != 0) {
  cat("The data is not complete.")
}
```

## GDP per capita per year

```{r}
gdp_capita_year <- 
  gdp_capita_year_url %>% 
  read_csv()

if (sum(is.na(gdp_capita_year)) != 0) {
  cat("The data is not complete.")
}
```

## Country names and continents

```{r}
geo <- 
  geo_url %>% 
  read_csv()

if (sum(is.na(geo)) != 0) {
  cat("The data is not complete.")
}

missing_continent <- 
  tribble(
    ~Entity, ~Group,
    "Central African Republic", "Africa",
    "Czech Republic", "Europe",
    "Dominican Republic", "America",
    "North Korea", "Asia",
    "South Korea", "Asia",
    "Kyrgyz Republic", "Asia",
    "Lao", "Asia",
    "St. Lucia", "America",
    "St. Vincent and the Grenadines", "America",
    "Yemen", "Asia",
    "South Sudan", "Africa")

continent <-
  googlesheets::gs_url(continent_url) %>% 
  googlesheets::gs_read(ws = 2) %>% 
  select(Entity:Group) %>% 
  mutate(Group =  Group %>% 
                    stringr::str_replace_all( "[\\[\\]]", "")) %>% 
  bind_rows(missing_continent)
  
  
```

## Population

```{r}
pop <- 
  population_url %>% 
  read_csv()
```


# Joining the data together

```{r}
data <- 
  life_expectancy %>% 
  inner_join(gdp_capita_year,
             by = c("geo", "time")) %>% 
  inner_join(geo, by = "geo") %>% 
  left_join(continent,
            by = c("name" = "Entity")) %>% 
  rename(continent = Group,
         country = name,
         year = time,
         gdp_capita = gdp_per_capita_cppp) %>% 
  left_join(pop,
            by = c("year" = "year", "geo" = "country"))

```

# Plot 
## Plot parameters

```{r}
continent_col <- 
  c("Africa" = "#00D5E9",
    "Asia" = "#FF5872",
    "America" = "#7FEB00",
    "Europe" = "#FFE700")
```



## Plotting without interaction

```{r}
plot_year <- function(year_) {
  data %>% 
    filter(year == year_) %>% 
    arrange(desc(population)) %>% 
    ggplot(aes(gdp_capita, life_expectancy)) +
    annotate("text", x = 7000, y = 50, label = year_,
             size = 30, colour = "#E0E0E0", family = "Courier") +
    geom_point(aes(size = population, fill = continent),
               shape = 21, colour = "grey25", show.legend = FALSE) +
    coord_cartesian(xlim = c(350, 500*2^8),
                    ylim = c(15, 88)) +
    scale_x_continuous(trans = "log2",
                       breaks = 500*2^(0:8),
                       labels = c(500*2^(0:4), "16k", "32k", "64k", "128k"), 
                       minor_breaks = NULL,
                       name = "Income per person") +
    scale_y_continuous(breaks = seq(20, 80, 10), minor_breaks = NULL,
                       name = "Life expectancy, years") +
    scale_size(limits = range(data$population), range = c(0.25,20)) +
    scale_fill_manual(values = continent_col) +
    theme_minimal() +
    theme(axis.line = element_line(color = "#999999", size = 0.25),
          panel.grid.major = element_line(linetype = "dotted", color = "#999999", size = 0.25),
          axis.text = element_text(family = "Courier", color = "#999999"),
          axis.title.x = element_text(hjust = 0, colour = "#425663"),
          axis.title.y = element_text(angle = 0, hjust = 0, vjust = 1.08, 
                                      margin = margin(0, -110, 0, 10), colour = "#425663"),
          plot.margin = unit(c(2.5, 1, 1.25, 1), "lines"))
  }
```

Year 1800 - 2000:
  
```{r}
seq(1800, 2000, 50) %>% 
  map(plot_year)
```


## Plot with interaction

### ggplot + plotly

```{r}
p <-
 data %>%
   #arrange(desc(population)) %>%
   ggplot(aes(gdp_capita, life_expectancy)) +
   # annotate("text", x = 7000, y = 50, label = year,
   #         size = 30, colour = "#E0E0E0", family = "Courier") +
   geom_point(aes(size = population, fill = continent, frame = year),
              shape = 21, colour = "grey25", show.legend = FALSE) +
   coord_cartesian(xlim = c(350, 500*2^8),
                   ylim = c(15, 88)) +
   scale_x_continuous(trans = "log2",
                      breaks = 500*2^(0:8),
                      labels = c(500*2^(0:4), "16k", "32k", "64k", "128k"),
                      minor_breaks = NULL,
                      name = "Income per person") +
   scale_y_continuous(breaks = seq(20, 80, 10), minor_breaks = NULL,
                      name = "Life expectancy, years") +
   scale_size(limits = range(data$population), range = c(0.25,20)) +
   scale_fill_manual(values = continent_col) +
   theme_minimal() +
   theme(axis.line = element_line(color = "#999999", size = 0.25),
         panel.grid.major = element_line(linetype = "dotted", color = "#999999", size = 0.25),
         axis.text = element_text(family = "Courier", color = "#999999"),
         axis.title.x = element_text(hjust = 0, colour = "#425663"),
         axis.title.y = element_text(angle = 0, hjust = 0, vjust = 1.08,
                                     margin = margin(0, -110, 0, 10), colour = "#425663"),
         plot.margin = unit(c(2.5, 1, 1.25, 1), "lines"))
```

```{r}
ggplotly(p,
         width = 900,
         height = 900*0.6) %>%
 animation_opts(200, easing = "linear-in-out")
```

Plotly doesn't play well with stylised ggplot. Let's see how closely we can recreate the Gapminder plot with Plotly only.

### Pure Plotly

```{r}
font_title <- 
  list(family = "Courier",
       size = 16,
       color = "#425663")

font_axis <- 
  list(family = "Courier",
       size = 12,
       color = "#999999")

slider_opts <- 
  list(prefix = "YEAR ",
       font = list(family = "Courier",
                   size = 20, 
                   color = "#425663"))

x_axis <- 
  list(title = "Income per person",
       type = "log",
       ticktext = c(500*2^(0:4), "16k", "32k", "64k", "128k"),
       tickvals = 500*2^(0:8),
       zeroline = FALSE,
       titlefont = font_title,
       tickfont = font_axis,
       range = c(log(300)/log(10), log(500*2^8.5)/log(10)),
       showline = TRUE,
       ticks = "outside",
       tickcolor = "white",
       ticklen = 10)

y_axis <- 
  list(title = "Life expectancy, years",
       tickvals = seq(20, 80, 10),
       range = c(15, 88),
       titlefont = font_title,
       tickfont = font_axis,
       showline = TRUE,
       ticks = "outside",
       tickcolor = "white",
       ticklen = 10)

data %>%
  plot_ly(x = ~gdp_capita, 
          y = ~life_expectancy,
          text = ~country, 
          hoverinfo = "text",
          size = ~population,
          sizes = c(5, 4000),
          width = 900,
          height = 900*0.6) %>%
  layout(xaxis = x_axis,
        yaxis = y_axis) %>%
  add_trace(color = ~continent, 
            colors = ~continent_col,
            frame = ~year, 
            ids = ~country,
            mode = "markers",
            marker = list(
                      opacity = 1,
                      line = list(
                              color = I("black"),
                              width = 1
                             )
                      ),
           type = "scatter") %>%
  animation_opts(frame = 0, 
                 easing = "linear-in-out") %>%
  animation_slider(currentvalue = slider_opts)
```

This is the closest to the orginal plot I could get.

* I couldn't get the `frame` animation system to work with both the marker and the in-plot year annotation.
* I cannot modify the grid line linetype.
* I couldn't decouple `size` and `color` for the legend.
* The `frame` animation system is undocumented: I don't know how / what can be tweaked in the design of the animation button / slider / axis.
* I don't know how to set the plotting order to show the smaller countries on top of the large ones.

